A Data-Driven Framework for Neural Field Modeling

### 2011-01-07 ltx/first_order_framework/LaTeX_Manuscript.tex
removed the word 'nice' from line 625. I'm worried we're getting a bit too chatty in this document!
edits to discussion

\dean{The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the Galerkin projection functions and the states would be the amplitude coefficients. If the Galerkin approach was used estimate the states, $n_x$ coupled ODEs would be solved independently to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation. To show how this is achieved we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving

@@

  }

@@

The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. \dean{Alternative field basis functions may be used to form the decomposition and if the basis is orthonormal then $\Gamma$ is simply the identity matrix.} Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where

@@

Now we define the state disturbance as
\begin{equation}\label{eq:Wt} 
	\mathbf{e}_t \triangleq \boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} ) \, d\mathbf{r}},
\end{equation}
which is a zero mean normally distributed white noise process with covariance (see \ref{App:ColoredNoise} for the derivation)
\begin{equation}
	\boldsymbol\Sigma_e =\mathbf{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top} \, d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top}. 
\end{equation}
The observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into equation~\ref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)\boldsymbol{\phi}^{\top}\left(\mathbf{r'}\right) \mathbf{x}_t\, d\mathbf{r}'} + \boldsymbol{\varepsilon}_t. 
\end{equation}
\dean{Note, the sensor index $\mathbf{r}_n$ has been omitted to delineate the reduced model observations from the full model.} The observation equation is linear and can be written in the more compact form
\begin{equation}\label{ObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix is 
\begin{equation}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\phi}_j(\mathbf{r}') \, d\mathbf{r}'.
\end{equation}

@@

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation or fixed-lag, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF \parham{and an online recursive LS estimator (see \citet{Ljung1999} for example)}. }

@@

\subsection{Extensions to the framework}
In this study, an assumption was made that the chosen kernel basis functions can exactly represent the actual connectivity kernel. In future work, we plan to relax this assumption. In addition, we have assumed that the neural field is isotropic and homogeneous. It is generally accepted that the intracortical connectivity structure can be accurately modelled as being isotropic. However, the assumption of homogeneity is quite strong as the size of the neural field becomes larger. Future work should focus on developing estimation algorithms where assumptions of homogeneity can be relaxed. In addition, natural extensions of this framework are the inclusion of a post-synaptic response kernel with a finite rise and decay time, and a finite distant-dependant action potential propagation velocity into the parametric neural field equations.

\section{Conclusion}
This paper has presented a framework for creating data-driven neural field models by developing a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. In doing so, we have highlighted important aspects of the experiment design process, demonstrating the relationship between sensor bandwidth and the complexity of the field that can be inferred. We have derived a finite-order state-space representation of the continuum neural field equations that allows inference to be performed efficiently. Importantly, the mathematical formulation that we have presented scales well with the number of sensors. The major contribution of this framework is to formally bridge a gap between experimental and theoretical neuroscience in order to gain a better understanding of the mesoscopic dynamics of the brain.

\section{Acknowledgments}
Dean Freestone would like to thank Dr. Mark van Rossum for hosting him while visiting the University of Edinburgh where this work was undertaken and Dr. Matthias Hennig for useful discussions and feedback. He would also like to thank Dr. James Fallon for assistance wit the observation kernel experiment. This research was funded in part by the Australian Research Council (Linkage Project LP0560684). The Bionic Ear Institute acknowledges the support it receives from the Victorian State Government through the Operational Infrastructure Support Program. Dean Freestone would like to thank the University of Melbourne's Scholarships Office for the Postgraduate Overseas Research Experience Scholarship and the Harold Mitchell Foundation for a traveling scholarship for supporting this research. Parham Aram would also like to thank Andrew Hills for useful discussions and assistance with code and the generation of figures. The authors would also like to thank Dr. John Terry for insightful discussions and feedback.

@@

\section{Equivalence to Galerkin's Projection}
To show how to solution of equation for the state vector is equivalent to the Galerkin projection we define the residual between the field and the decomposition as 
\begin{align}
	\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) &= v_{t+1}(\mathbf{r}) - T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection method finds $\mathbf{x}_t$, such that for all of the basis functions, the inner product with the residual satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0, \> i=0,1,\hdots,n_x-1,
\end{equation}
where $n_x$ is the number of states and basis functions. If the inner product between the residual and the basis functions in equation~\ref{eq:GalerkinProjection} is zero, then the basis functions are orthogonal to the residual.

>>>>

\dean{The solution of this equation for $\mathbf{x}_t$ is equivalent to finding a Galerkin projection, where the field basis functions would be considered the Galerkin projection functions and the states would be the amplitude coefficients. In the next step to forming the state-space model we substitute the field decomposition into the left hand side of equation~\ref{eq:decompRHS} giving}

@@

@@

The analytic derivation of the inner product of \emph{n}-dimensional Gaussians, which was used for calculating $\mathbf\Gamma$, is provided in \ref{App:InnerProdOfGaussians}. \dean{Alternative field basis functions may be used to form the decomposition and if the basis is orthonormal, such as a Fourier basis, then $\Gamma$ is simply the identity matrix.} Equation~\ref{eq:ReducedForm} can be simplified by exploiting the isotropy of the connectivity kernel basis functions where

@@

Now we define the state disturbance as
\begin{equation}\label{eq:Wt} 
	\mathbf{e}_t \triangleq \boldsymbol{\Gamma}^{-1}\int_\Omega {\boldsymbol{\phi} ( \mathbf{r} )e_t( \mathbf{r} ) \, d\mathbf{r}},
\end{equation}
which is a zero mean normally distributed white noise process with covariance (see \ref{App:ColoredNoise} for the derivation)
\begin{equation}
	\boldsymbol\Sigma_e =\mathbf{\Gamma}^{-1}\int_{\Omega}\int_{\Omega}\boldsymbol{\phi}\left(\mathbf r\right) \gamma\left(\mathbf r- \mathbf r' \right)\boldsymbol{\phi}\left(\mathbf r'\right)^{\top} \, d\mathbf r' d\mathbf r\mathbf{\Gamma}^{- \top}. 
\end{equation}
The observation equation of the reduced model is found by substituting equation~\ref{DefFieldDecomp} into equation~\ref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation}
	\mathbf{y}_t = \int_{\Omega}{m\left(\mathbf{r}_n-\mathbf{r}'\right)\boldsymbol{\phi}^{\top}\left(\mathbf{r'}\right) \mathbf{x}_t\, d\mathbf{r}'} + \boldsymbol{\varepsilon}_t. 
\end{equation}
\dean{Note, the sensor index $\mathbf{r}_n$ has been omitted to delineate the reduced model observation equation from the full model.} The observation equation is linear and can be written in the more compact form
\begin{equation}\label{ObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\varepsilon}_t,
\end{equation}
where each element of the observation matrix is 
\begin{equation}
	\mathbf{C}_{ij} \triangleq \int_{\Omega}m(\mathbf{r}_i - \mathbf{r}')\boldsymbol{\phi}_j(\mathbf{r}') \, d\mathbf{r}'.
\end{equation}

@@

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based or fixed-lag estimation, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF \parham{and an online recursive LS estimator (see \citet{Ljung1999} for example)}. }

@@

\subsection{Extensions to the framework}
In this study, an assumption was made that the chosen kernel basis functions can exactly represent the actual connectivity kernel. In future work, we plan to relax this assumption \dean{and show how the support of the connectivity kernel can also be inferred from data}. In addition, we have assumed that the neural field is isotropic and homogeneous. It is generally accepted that the intracortical connectivity structure can be accurately modelled as being isotropic. However, the assumption of homogeneity is quite strong as the size of the neural field becomes larger. Future work should focus on developing estimation algorithms where assumptions of homogeneity can be relaxed. In addition, natural extensions of this framework are the inclusion of a post-synaptic response kernel with a finite rise and decay time, and a finite distant-dependant action potential propagation velocity into the parametric neural field equations.

\section{Conclusion}
This paper has presented a framework for creating data-driven neural field models by developing a method for estimating intracortical connectivity and synaptic dynamics from electrophysiological data. In doing so, we have highlighted important aspects of the experiment design process, demonstrating the relationship between sensor bandwidth and the complexity of the field that can be inferred. We have derived a finite-order state-space representation of the continuum neural field equations that allows inference to be performed efficiently. Importantly, the mathematical formulation that we have presented scales well with the number of sensors. The major contribution of this framework is to formally bridge a gap between experimental and theoretical neuroscience in order to gain a better understanding of the mesoscopic dynamics of the brain. \dean{The validation of the estimation framework on real data is now required, and is the subject of planned future work.}

\section{Acknowledgments}
Dean Freestone would like to thank Dr. Mark van Rossum for hosting him while visiting the University of Edinburgh where this work was undertaken and Dr. Matthias Hennig for useful discussions and feedback. He would also like to thank Dr. James Fallon for assistance wit the observation kernel experiment. This research was funded in part by the Australian Research Council (Linkage Project LP0560684). The Bionic Ear Institute acknowledges the support it receives from the Victorian State Government through the Operational Infrastructure Support Program. Dean Freestone would like to thank the University of Melbourne's Scholarships Office for the Postgraduate Overseas Research Experience Scholarship and the Harold Mitchell Foundation for a traveling scholarship for supporting this research. Parham Aram would also like to thank Andrew Hills for useful discussions and assistance with code and the generation of figures. The authors would also like to thank Dr. John Terry for insightful discussions and feedback. \dean{Finally, the authors would like to thanks the anonymous reviews who have provided an excellent critique that has led to valuable insights that would have gone otherwise astray.}

@@

\section{Equivalence to Galerkin's Projection}
\dean{To show how to solution of equation for the state vector is equivalent to the Galerkin projection we define the residual between the field and the decomposition as 
\begin{align}\label{eq:BasisFunctionResidual}
	\Xi(\mathbf{x}_t,v_{t+1}&(\mathbf{r})) = v_{t+1}(\mathbf{r}) \nonumber \\
	&- T_s\int_\Omega{f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t )\boldsymbol{\psi}^{\top}(\mathbf{r}-\mathbf{r}') \, d\mathbf{r}'}\boldsymbol{\theta} \nonumber \\
	&+ \xi\boldsymbol{\phi}^{\top}(\mathbf{r})\mathbf{x}_t + \mathbf{e}_t(\mathbf{r}).
\end{align} 
The Galerkin projection method finds $\mathbf{x}_t$, such that the inner product of the residual with all of the basis functions satisfies
\begin{equation}\label{eq:GalerkinProjection}
	G_i = \int_{\Omega}\Xi(\mathbf{x}_t,v_{t+1}(\mathbf{r})) \phi_i(\mathbf{r}) d\mathbf{r} = 0,
\end{equation}
where $i=0,1,\hdots,n_x-1,$ and $n_x$ is the number of states and basis functions. If the inner product between the residual and the basis functions in equation~\ref{eq:GalerkinProjection} is zero, then the basis functions are orthogonal to the residual. By substituting equation~\ref{eq:BasisFunctionResidual} into equation~\ref{eq:GalerkinProjection} and rearranging we get equation~\ref{eq:decompRHS} in the main text. Note, if Galerkin's method was used estimate the states, $n_x$ coupled ODEs would need to be solved independently (each $G_i$) to estimate the state vector, which is a standard method in fluid dynamics and economics modeling. The method in this paper takes a different approach by forming the state-space model, allowing for more efficient estimation. }
